---
title: "Final Project"
subtitle: "Proposal"
author: 
  - name: "Pascal"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Comparing the box office, profitability, and ratings of films based on video games."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
game_films <- read_csv("data/game_films.csv")

glimpse(game_films)
```

```{r}
#| label: dataset-scope
game_films |> 
  count(category, sort = TRUE)
```

```{r}
#| label: proposal-counts
#| include: false
n_usd <- sum(game_films$worldwide_box_office_currency == "$", na.rm = TRUE)
n_profit <- game_films |>
  filter(!is.na(worldwide_box_office), !is.na(budget_low)) |>
  nrow()
```

This TidyTuesday dataset, **Films Based on Video Games**, is scraped from the Wikipedia article [*List of films based on video games*](https://en.wikipedia.org/wiki/List_of_films_based_on_video_games) and curated by Georgios Karamanis. Each row is one film adaptation, with `r ncol(game_films)` variables spanning `r nrow(game_films)` films released since the early 1980s to upcoming titles.

Source: [TidyTuesday, 2026-06-09](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-06-09), scraped from Wikipedia's *List of films based on video games*.

It covers theatrical releases, direct-to-video, TV films, shorts, and documentaries, recording box office, budgets, critic and audience scores, distributors, and the original game publisher. I chose it because game-to-film adaptations have been growing greatly in the past decades, which raises questions as to how these adaptations actually do.

Question 2 will additionally merge an external video-games ratings dataset (more on that below).

## Questions

1.  **Commercial performance.** Among films adapted from video games, how does worldwide box office vary based on the publisher of the video game that was adapted into a film and the time in which it was released, and how often do these films even break even?

2.  **Game vs. film ratings.** For video game franchises adapted into film, how do the films' review scores compare to the ratings of their corresponding original games, and does this game-vs-film gap differ across publishers or different times?

## Analysis plan

### Question 1: Commercial performance

- **Variables used:** `worldwide_box_office`, `worldwide_box_office_currency`, `budget_low`, `budget_high`, `original_game_publisher`, `release_date`.

- **Variables to create:** `release_era` (decade, from `release_date`), `profit` and `roi` (box office to mid-budget), `box_office_real` (adjusting for inflation via CPI), a collapsed publisher factor (top publishers + "Other").

- **Filtering:** restrict to `$(USD)` box office for comparability (`r n_usd` films); box office + budget overlap is `r n_profit` films, which I will use as the profitability sample.

- **Approach:** box office and return on investment summarized by publisher (bar/boxplot) and by era (trend), plus a budget vs box-office scatter plot to show which adaptations actually turned a profit.

- **External Data merged:** *Annual U.S. CPI*, used to convert box office amounts into inflation-adjusted dollars. 

### Question 2: Game vs. film ratings

- **Variables used (films):** `metacritic`, `rotten_tomatoes`, `cinema_score`, `original_game_publisher`, `release_date`.

- **Variables to create:** a `franchise` key applied to both tables via `str_detect` (e.g., all "Resident Evil" titles -\> one franchise), franchise level averages of game and film scores, a `rating_gap` = average game score - average film score, numeric conversion of `cinema_score` letter grades for an audience comparison.

- **External Data merged:** *Video Game Sales with Ratings* (VGChartz + Metacritic), providing `Critic_Score` (0-100, same scale as film Metacritic) and `User_Score` (0-10) per game.

- **Join:** `group_by(franchise) |> summarize(...)` on each side, then `left_join`, giving one row per franchise.

- **Approach:** a scatter plot of the average game score vs. the average film score per franchise, with a reference line. Points above the line are franchises whose films were better-reviewed than their games, as opposed to points below which are the opposite. I will then group the `rating_gap` (game score - film score) by publisher and by era to see whether certain publishers or periods show consistently larger gaps.

## Plan of attack

*Living document - updated at least weekly.*

| Task Name | Status | Assignee | Due | Priority | Summary |
|------------|------------|------------|------------|------------|------------|
| Proposal & questions | Done | Pascal | Week 3 | High | Finalize dataset, two questions, analysis plan |
| Wrangle film data | started | Pascal | Week 4 | High | Parse release era, filter USD box office, create profit/roi |
| Review feedback | Not started | Pascal | Week 5 | Medium | Implement user feedback |
| Write-up | Not started | Pascal | Week 6 | High | Create plots and finalize Analysis |
| Presentation | Not started | Pascal | Week 7 | Medium | Get slides ready for presentation |